Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2012-12-02 2.tex
First complete draft!

One of the major  for the development of stochastic models o
motivations for the 
Recent developments in the theory of phonotactic knowledge are motivated by the claim that speakers' wordlikeness intuitions are inherently \emph{gradient}, i.e., that such intuitions are more granular than implied by a binary contrast between ``possible'' and ``impossible'' words.

@@

Scientific observations do not necessarily arrive at the appropriate granularity for analysis.
Since the granularity of measurement and analysis is determined by prior hypotheses about the nature of the system under study, disputes are inevitable as scientific paradigms evolve. 
This study argues that there are both theoretical and empirical reasons to doubt the implicit hypothesis linking scalar wordlikeness judgements and a gradient model of phonotactics. 
First, intermediate judgements are characteristic of all gradient rating tasks, and therefore these judgements are irrelevant to the question of whether well-formedness is categorical or gradient.
Secondly, simple baselines better account for gradient well-formedness judgements than current computational models of phonotactic knowledge, suggesting that the gradience observed in these tasks do not derive from grammatical mechanisms.

@@

As first noted by \citet{Chomsky1963b}, speakers experience difficulty processing sentences with multiple center embeddings. \citet{Gibson1999} find that speakers rate sentences like (\ref{gibson}a), which is well-formed, less grammatical than (\ref{gibson}b), which is nonsensical.

@@

\noindent
It is informative to consider that this well-known result has had no effect on the theory of syntactic representations, only on the theory of linguistic memory; it is recognized as the product of cognitive restrictions found in non-linguistic domains, as a \emph{task effect}.
This contrasts with the argument made by \citet{Hayes2000}, that gradient wordlikeness judgements demand an all-encompassing revision to the grammatical architecture, reviewed below.

@@

\noindent
However, when subjects are asked to rate, using a 7-point Likert scale, how representative individual odd counting numbers are of the concept \emph{odd number}, they freely use intermediate ratings; their results with \emph{odd number} and \emph{even number} are shown in Figure \ref{agg}.

\begin{figure}[ht]
\centering
\includegraphics{agg.pdf}
\caption{Subjects freely use intermediate ratings when asked to rate how representative even and odd numbers were of ``even'' and ``odd'', respectively (\citealp{Armstrong1983}; ``1'' indicates ``most representative'')}
\label{agg}
\end{figure}

@@

\noindent
This is indisputable. 
However, there have been no serious attempts to evaluate categorical and gradient models of wordlikeness on an equal footing.
In light of the complexities of gradient models, such an evaluation requires strong quantitative evidence for the superiority of gradient grammatical models.
This study represents a first attempt to fill this gap.

@@

\begin{table}[ht]
\centering
\begin{tabular}{l rrr}
\toprule
                           & subjects & items & trials \\
\midrule
\citeauthor{Albright2007}  & 68       & 40    & 2,720  \\
\citeauthor{Albright2003b} & 24       & 86    & 2,064  \\
\citeauthor{Scholes1966}   & 33       & 63    & 2,178  \\
\midrule
\textsc{Total}             & 125      & 187   & 6,962  \\
\bottomrule
\end{tabular}
\caption{Subject and item counts}
\label{counts}
\end{table}

@@

\citet{Albright2003b} have 24 adult speakers rate 87 aurally presented monosyllabic nonce words on a 7-point Likert scale with endpoints labeled ``completely bizarre, impossible as an English word'' and ``completely normal, would make a fine English word''. This task was administered to establish phonotactic norms for a later nonce word inflection task. Their item [raɪf] is excluded in this study, since this is an actual word of English, \emph{rife}. \citet{Albright2009a} uses this data to evaluate several computational models of wordlikeness.

@@

\citet{Scholes1966} conducts several wordlikeness tasks with 7th-grade students (approximately 12--13 years of age). The data used here is his experiment 5, in which 33 speakers provide a ``yes'' or ``no'' as to whether each of the 63 items, presented aurally, are ``likely to be usable as a word of English''. Like the study by \citet{Albright2007}, the focus is on onset well-formedness and there is minimal diversity in rime type. Two items, [klʌŋ] \emph{clung} and [bɹʌŋ] \emph{brung} (a dialectal past participle of \emph{bring}), are excluded here as actual words of English. \citet{Albright2009a} and \citet{Hayes2008a} also use this data for the purposes of model evaluation; following \citet{Frisch2000}, they use the proportion of ``yes'' responses for each item so as to derive a continuous measure of well-formedness.

@@

Much easier to interpret are two other non-parametric statistics, the Goodman-Kruskal $\gamma$ and the Kendall $\tau_b$ \citep{Noether1981}, as follows. These statistics are computed by comparing every model score/wordlikeness rating pair to every other: a comparison is counted as \emph{concordant} if the greater of the two model scores is the one associated with the greater of the two wordlikeness ratings (that is, the model ranks these two nonce words in accordance with speakers' ratings), and as \emph{discordant} otherwise. These two statistics differ only in the treatment of ``ties'', pairs where either the model score or wordlikeness rating are identical. For $\gamma$, ties are ignored, and the coefficient is

@@

The nonce word stimuli from these three studies are scored automatically using four computational models. The first two models represent baselines for comparison to the latter two state-of-the-art gradient models. The scores for each model are provided in Appendix \ref{ratings}.

@@

\begin{figure}[ht]
\centering
\includegraphics{boxplot.pdf}
\caption{Gross phonotactic status and item-averaged wordlikeness ratings}
\label{boxplot}
\end{figure}

@@

A wide variety of lexical similarity measures were considered, including a variant of the Generalized Neighborhood model \citep{Bailey2001}, PLD20 \citep{Suarez2011}, and a set of measures provided by the Irvine Phonotactic Calculator \citep{Vaden2009}. The most reliable measure is also the most venerable measure of lexical similarity: Coltheart's $N$ \citep{Coltheart1977}, which is defined as the number of words in some representative sample which can be changed into a target nonce word by a single insertion, deletion, or substitution of a phone. \citet{Greenberg1964} find a correlation between wordlikeness ratings and a variant of this measure which only counts words differing by a single substitution. This measure is plotted against ratings from the three studies in Figure \ref{neighborhood}, where it can be seen that it accounts for much of the variance in ratings.

@@

\begin{figure}[ht]
\centering
\includegraphics{neighborhood.pdf}
\caption{Correlation between neighborhood density and item-averaged wordlikeness ratings}
\label{neighborhood}
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{l rrrr}
\toprule
                                 & Pearson $r$ & Spearman $\rho$ & G-K $\gamma$ & Kendall $\tau_{b}$ \\
\midrule
featural  bigrams                &       {.71} &       {.64}     &       {.45}  &       {.45} \\
segmental bigrams                &       {.74} &       {.67}     &       {.48}  &       {.47} \\
segmental bigrams with smoothing & \uline{.75} & \uline{.70}     & \uline{.50}  & \uline{.50} \\
\bottomrule
\end{tabular}
\caption{Correlation between item-averaged wordlikeness ratings for the \citet{Albright2003b} norming study and three variants of bigram probability}
\label{bigramcomparison}
\end{table}

\begin{figure}[ht]
\centering
\includegraphics{bigram.pdf}
\caption{Correlation between segmental bigram score and item-averaged wordlikeness ratings}
\label{bigram}
\end{figure}

@@

Since this model has numerous experimenter-defined parameters, a close replication of \citeauthor{Hayes2008a}'s original paper is attempted: both their implementation and phonological feature specifications are used here.
Following \citet{HayesInPress}, dictionary entries are syllabified using the procedure described in \S\ref{syllabification} and a novel feature [$\pm$\textsc{Coda}] is added to allow the model to distinguish coda and onset consonants.
Also, following \citeauthor{Hayes2008a}, constraints are limited to those spanning as many as three segments and an ``accuracy schedule'' of [$.001$, $.01$, $.1$, $.2$, $.3$] is used.
Since the maximum entropy model produces slightly different scores on each run, the worst-performing of 10 runs is reported here, following \citeauthor{Hayes2008a}.
The resulting scores are plotted against wordlikeness ratings in Figure \ref{maxent}; it can be seen that the model assigns the highest possible score to a large variety of nonce words, though many of these words appear to have a low rating. This suggests that the model is not robust enough to score whole words.

\begin{figure}[ht]
\centering
\includegraphics{maxent.pdf}
\caption{Correlation between MaxEnt score and item-averaged wordlikeness ratings}
\label{maxent}
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{l rrr c rrr c rrr c rrr}
\toprule
              &         \multicolumn{3}{c}{Pearson $r$} &&     \multicolumn{3}{c}{Spearman $\rho$} && \multicolumn{3}{c}{G-K $\gamma$} && \multicolumn{3}{c}{Kendall $\tau_b$} \\
\cmidrule{2-4} \cmidrule{6-8} \cmidrule{10-12} \cmidrule{14-16}
              &           A &         AH &            S &&           A &       AH    &           S &&           A &          AH &           S &&           A &          AH & S \\
\midrule
Gross status & \uline{.73} &       {.60} &       {.80} && \uline{.82} &       {.66} &       {.80} && \uline{.87} & \uline{.93} & \uline{.91} &&       {.67} &       {.47} &       {.62} \\
Density ($N$) &       {.67} & \uline{.79} & \uline{.86} &&       {.61} & \uline{.74} & \uline{.82} &&       {.49} &       {.57} &       {.74} &&       {.45} & \uline{.56} & \uline{.67} \\
Bigram $p$ &       {.46} &       {.75} &       {.74} &&       {.34} &       {.70} &       {.79} &&       {.25} &       {.50} &       {.63} &&       {.25} &       {.50} &       {.61} \\
MaxEnt $p$ &       {.70} &       {.21} &       {.53} &&       {.66} &       {.39} &       {.58} &&       {.85} &       {.61} &       {.56} && \uline{.68} &       {.16} &       {.48} \\ 
\bottomrule
\end{tabular}
\caption{Correlations between item-averaged wordlikeness ratings and model scores; the largest coefficient for each statistic/data set pair is underlined.}
\label{scores}
\end{table}

@@

\begin{table}[ht]
\centering
\begin{tabular}{l rrr c rrr}
\toprule                
             & \multicolumn{3}{c}{``Valid'' items} && \multicolumn{3}{c}{``Invalid'' items} \\
\cmidrule{2-4} \cmidrule{6-8}
             %& \multicolumn{3}{c}{Kendall $\tau_{b}$} && \multicolumn{3}{c}{Kendall $\tau_{b}$} \\
             &         A &       AH &        S &&     A  &       AH &          S \\
\midrule
Bigram score &     {.65} &    {.34} &    {.60} &&    {.03} & {$-$.17} &    {.47} \\
MaxEnt score &     {.00} & {$-$.15} & {$-$.32} && {$-$.42} &    {.29} & {$-$.16} \\
\bottomrule
\end{tabular}
\caption{Kendall $\tau_{b}$ correlation between item-averaged wordlikeness ratings (rated ``valid'' or ``invalid'' by the gross phonotactic status baseline) and model scores}
\label{resid}
\end{table}

@@

The bigram and MaxEnt models do not reliably outperform simple baselines. From this it can be inferred that the gradient models do not reliably predict intermediate ratings. These models do not reliably distinguish within classes of ``valid'' and ``invalid'' words either.

>>>>

One of the major  for the development of stochastic models o
motivations for the 
Recent developments in the theory of phonotactic knowledge are motivated by the claim that speakers' wordlikeness intuitions are inherently \emph{gradient}, i.e., that such intuitions are more granular than a mere binary contrast between ``possible'' and ``impossible'' words.

@@

Scientific observations do not necessarily arrive at the appropriate granularity for analysis.
Since the granularity of measurement and analysis is determined by prior hypotheses about the nature of the system under study, disputes are inevitable as scientific paradigms evolve. 
This chapter argues that there are both theoretical and empirical reasons to doubt the implicit hypothesis linking scalar wordlikeness judgements and a gradient model of phonotactics. 
First, intermediate judgements are characteristic of all gradient rating tasks, and therefore these judgements are irrelevant to the question of whether well-formedness is categorical or gradient.
Secondly, simple baselines better account for gradient well-formedness judgements than current computational models of phonotactic knowledge, suggesting that the gradience observed in these tasks do not derive from grammatical mechanisms.

@@

As first noted by \citet{Chomsky1963b}, speakers experience difficulty processing sentences with multiple center embeddings. 
\citet{Gibson1999} perform a controlled experiment which reveals that speakers rate sentences like (\ref{gibson}a), which is well-formed, less grammatical than (\ref{gibson}b), despite the fact that a moment of reflection reveals that the latter sentence is nonsensical.

@@

\noindent
It is informative to consider that this well-known result has had no effect on the theory of syntactic representations, only on the theory of linguistic memory; it is recognized as the product of cognitive restrictions found in non-linguistic domains, as a \emph{task effect}.
This contrasts with the argument made by \citet{Hayes2000}, that gradient wordlikeness judgements demand a considerable and unmotivated revision to the grammatical architecture, as is discussed below.

@@

\noindent
However, when subjects are asked to rate, using a 7-point Likert scale, how representative individual odd counting numbers are of the concept \emph{odd number}, they freely use intermediate ratings; the ratings they obtain with instances of \emph{odd number} and \emph{even number} are shown in Figure \ref{agg}.

\begin{figure}[t]
\centering
\includegraphics{agg.pdf}
\caption{Subjects freely use intermediate ratings when asked to rate how representative even and odd numbers were of ``even'' and ``odd'', respectively (\citealp{Armstrong1983}; ``1'' indicates ``most representative'')}
\label{agg}
\end{figure}

@@

\noindent
This is indisputable. 
However, there have been no serious attempts to evaluate categorical and gradient models of wordlikeness on an equal footing.
In light of the complexities of gradient models, such an evaluation requires strong quantitative evidence for the superiority of gradient grammatical models.
The evaluation below represents a first attempt to fill this gap.

@@

\begin{table}[t]
\centering
\begin{tabular}{l rrr}
\toprule
                           & subjects & items & trials \\
\midrule
\citeauthor{Albright2007}  & 68       & 40    & 2,720  \\
\citeauthor{Albright2003b} & 24       & 86    & 2,064  \\
\citeauthor{Scholes1966}   & 33       & 63    & 2,178  \\
\midrule
\textsc{Total}             & 125      & 187   & 6,962  \\
\bottomrule
\end{tabular}
\caption{Subject and item counts for the wordlikeness study}
\label{counts}
\end{table}

@@

\citet{Albright2003b} have 24 adult speakers rate 87 aurally presented monosyllabic nonce words on a 7-point Likert scale with endpoints labeled ``completely bizarre, impossible as an English word'' and ``completely normal, would make a fine English word''. This task was administered to establish phonotactic norms for a later nonce word inflection task. Their item [raɪf] is excluded in this study, since this is an actual word of English, \emph{rife}. \citet{Albright2009a} uses this data to compare computational models of wordlikeness.

@@

\citet{Scholes1966} conducts several wordlikeness tasks with 
students in 7th grade (approximately 12--13 years of age).
The data used here is his experiment 5, in which 33 speakers provide a ``yes'' or ``no'' as to whether each of the 63 items, presented aurally, are ``likely to be usable as a word of English''. 
Like the study by \citet{Albright2007}, the focus is on onset well-formedness and there is minimal diversity in rime type. 
Two items, [klʌŋ] \emph{clung} and [bɹʌŋ] \emph{brung} (a dialectal past participle of \emph{bring}), are excluded here as actual words of English. 
\citet{Albright2009a} and \citet{Hayes2008a} also use this data for the purposes of model evaluation; following \citet{Frisch2000}, they use the proportion of ``yes'' responses for each item so as to derive a continuous measure of well-formedness.

@@

Two other non-parametric statistics the Goodman-Kruskal $\gamma$ and the Kendall $\tau_b$ are much easier to interpret, as follows \citep{Noether1981}. 
These statistics are computed by comparing every model score/wordlikeness rating pair to every other: a comparison is counted as \emph{concordant} if the greater of the two model scores is the one associated with the greater of the two wordlikeness ratings (that is, the model ranks these two nonce words in accordance with speakers' ratings), and as \emph{discordant} otherwise. 
These two statistics differ only in the treatment of ``ties'', pairs where either the model score or wordlikeness rating are identical. 
For $\gamma$, ties are ignored, and the coefficient is

@@

The nonce word stimuli from these three studies are scored automatically using four computational models. The first two models represent baselines for comparison to the latter two state-of-the-art gradient models. The scores are reproduced in Appendix \ref{ratings}.

@@

\begin{figure}[t]
\centering
\includegraphics{boxplot.pdf}
\caption{Gross phonotactic status and item-averaged wordlikeness ratings}
\label{boxplot}
\end{figure}

@@

A wide variety of lexical similarity measures were considered, including a variant of the Generalized Neighborhood model \citep{Bailey2001}, PLD20 \citep{Suarez2011}, and a set of measures provided by the Irvine Phonotactic Calculator \citep{Vaden2009}. The most reliable measure is also the most venerable measure of lexical similarity: Coltheart's $N$ \citep{Coltheart1977}, which is defined as the number of words in some representative sample which can be changed into a target nonce word by a single insertion, deletion, or substitution of a phone. \citet{Greenberg1964} find a correlation between wordlikeness ratings and a variant of this measure which only counts words differing by a single substitution. This is plotted against ratings from the three studies in Figure \ref{neighborhood}, where it can be seen that it accounts for much of the variance in ratings.

@@

\begin{figure}[t]
\centering
\includegraphics{neighborhood.pdf}
\caption{Correlation between Coltheart's $N$ and item-averaged wordlikeness ratings}
\label{neighborhood}
\end{figure}

@@

\begin{table}[t]
\centering
\begin{tabular}{l rrrr}
\toprule
                                 & Pearson $r$ & Spearman $\rho$ & G-K $\gamma$ & Kendall $\tau_{b}$ \\
\midrule
featural  bigrams                &       {.71} &       {.64}     &       {.45}  &       {.45} \\
segmental bigrams                &       {.74} &       {.67}     &       {.48}  &       {.47} \\
segmental bigrams with smoothing & \uline{.75} & \uline{.70}     & \uline{.50}  & \uline{.50} \\
\bottomrule
\end{tabular}
\caption{Correlation between item-averaged wordlikeness ratings for the \citet{Albright2003b} norming study and three variants of bigram probability}
\label{bigramcomparison}
\end{table}

\begin{figure}[t]
\centering
\includegraphics{bigram.pdf}
\caption{Correlation between smoothed segmental bigram score and item-averaged wordlikeness ratings}
\label{bigram}
\end{figure}

@@

Since this model has numerous experimenter-defined parameters, a close replication of \citeauthor{Hayes2008a}'s original paper is attempted: both their implementation and phonological feature specifications are used here.
Following \citet{HayesInPress}, dictionary entries are syllabified using the procedure described in \S\ref{syllabification} and a novel feature [$\pm$\textsc{Coda}] is added to allow the model to distinguish coda and onset consonants.
Also, following \citeauthor{Hayes2008a}, constraints are limited to those spanning as many as three segments and an ``accuracy schedule'' of [$.001$, $.01$, $.1$, $.2$, $.3$] is used.
Since the maximum entropy model produces slightly different scores on each run, the worst-performing of 10 runs is reported here, following \citeauthor{Hayes2008a}.
The resulting scores are plotted against wordlikeness ratings in Figure \ref{maxent}; it can be seen that the model assigns the highest possible score to a large variety of nonce words, though many of these words appear to have a low rating. 
Apparently, this model is still not robust enough for extracting phonotactic generalizations from whole words.

\begin{figure}[t]
\centering
\includegraphics{maxent.pdf}
\caption{Correlation between MaxEnt score and item-averaged wordlikeness ratings}
\label{maxent}
\end{figure}

@@

\begin{table}[t]
\centering
\begin{tabular}{l rrr c rrr c rrr c rrr}
\toprule
              &         \multicolumn{3}{c}{Pearson $r$} &&     \multicolumn{3}{c}{Spearman $\rho$} && \multicolumn{3}{c}{G-K $\gamma$} && \multicolumn{3}{c}{Kendall $\tau_b$} \\
\cmidrule{2-4} \cmidrule{6-8} \cmidrule{10-12} \cmidrule{14-16}
              &           A &         AH &            S &&           A &       AH    &           S &&           A &          AH &           S &&           A &          AH & S \\
\midrule
Gross status & \uline{.73} &       {.60} &       {.80} && \uline{.82} &       {.66} &       {.80} && \uline{.87} & \uline{.93} & \uline{.91} &&       {.67} &       {.47} &       {.62} \\
Density ($N$) &       {.67} & \uline{.79} & \uline{.86} &&       {.61} & \uline{.74} & \uline{.82} &&       {.49} &       {.57} &       {.74} &&       {.45} & \uline{.56} & \uline{.67} \\
Bigram $p$ &       {.46} &       {.75} &       {.74} &&       {.34} &       {.70} &       {.79} &&       {.25} &       {.50} &       {.63} &&       {.25} &       {.50} &       {.61} \\
MaxEnt $p$ &       {.70} &       {.21} &       {.53} &&       {.66} &       {.39} &       {.58} &&       {.85} &       {.61} &       {.56} && \uline{.68} &       {.16} &       {.48} \\ 
\bottomrule
\end{tabular}
\caption{Correlation between item-averaged wordlikeness ratings and model scores}
%; the largest coefficient for each statistic/data set pair is underlined.}
\label{scores}
\end{table}

@@

\begin{table}[t]
\centering
\begin{tabular}{l rrr c rrr}
\toprule                
             & \multicolumn{3}{c}{``Valid'' items} && \multicolumn{3}{c}{``Invalid'' items} \\
\cmidrule{2-4} \cmidrule{6-8}
             %& \multicolumn{3}{c}{Kendall $\tau_{b}$} && \multicolumn{3}{c}{Kendall $\tau_{b}$} \\
             &         A &       AH &        S &&     A  &       AH &          S \\
\midrule
Bigram score &     {.65} &    {.34} &    {.60} &&    {.03} & {$-$.17} &    {.47} \\
MaxEnt score &     {.00} & {$-$.15} & {$-$.32} && {$-$.42} &    {.29} & {$-$.16} \\
\bottomrule
\end{tabular}
\caption{Kendall $\tau_{b}$ correlation between item-averaged wordlikeness ratings (rated the gross phonotactic status baseline) and model scores}
\label{resid}
\end{table}

@@

The bigram and MaxEnt models do not reliably outperform simple baselines. From this it can be inferred that the gradient models do not reliably predict intermediate ratings. Nor do these models reliably distinguish within classes of ``valid'' and ``invalid'' words in a way that conforms with wordlikeness ratings.
